Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-25 paper.tex

\subsection{Extensibility: GPU acceleration}
[JK: this is a super rough first narrative draft.]
Machine learning, and in particular deep learning methods, have led to
major breakthroughs in many different fields in recent years [CITES].
GPUs have been a key enabling factor in this [details and cites].
The PyData ecosystem [cites] has also enabled this. Techologies like
[Cupy, lots of others] substantially lower the bar for writing efficient
GPU code.
Deep learning methods depend crucially on efficient access to vast
datasets for training, and Zarr is emerging as a popular choice
[cites from geosciences?]. The RAPIDs GPU Accelerated data science
library [which does] now supports Zarr through [kvikio, more?].

@@

We first illustrate the ease with which computations can be
be peformed on the GPU, with little or no preprocessing of the
data. To do this we reimplemented the af-dist calculation
described above using CuPy~\citep{cupy}, a GPU accelerated library
of core operations for scientific computing.
All of the operations required to compute  af-dist are
performed on the GPU, mostly with minimal changes required
(e.g.\ replacing \texttt{np.digitize} with \texttt{cp.digitize}).
We perform the core operation of genotype classification using the
CuPy JIT compiler to generate a custom CUDA kernel, which
is also very similar to the Numba compiled function used in other
benchmarks. In the simplest version, we iterate over the chunks
in the \texttt{call\_genotype} array (stored as 30 chunks, consuming
a total of 7.82MiB), decompress each to a local
NumPy array and then copying the data to GPU memory by creating
the corresponding CuPy array. For this version, the total time
in the genotype classification kernel was 25ms with an
overall data transfer time of 179ms (we discount decompression
time here; see next paragraph). For comparison,
our standard Numba compiled classification function required
a total of 1.46s running on a single CPU core.
Another approach in which we load the entire
\texttt{call\_genotype} array into GPU memory at once (possible
here as the $96514 \times 2504 \times 2$ array requires
460MiB) and uses
a more sophisticated kernel in which we map GPU resources to
subsets of the array in two dimensions,
reduces transfer
time to 138ms and processing time to 6ms.

>>>>

\subsection{Extensibility: GPU acceleration}
GPU acceleration has revolutionized scientific computing by leveraging the
massively parallel architecture of GPUs to achieve significant speedups in
computationally intensive tasks such as simulations, data analysis, and machine
learning, often by one or two orders of magnitude~\citep{nickolls2010gpu,
mittal2014survey,owens2008gpu}
% (Nickolls & Dally, 2010;
% Mittal & Vetter, 2015; 
% Owens et al., 2008). 
This has profound implications for
fields ranging from molecular dynamics~\citep{stone2007accelerating}
% (Stone et al., 2010) 
to deep learning~\citep{lecun2015deep}
% (LeCun et al., 2015). 
While early GPU programming required expertise in
low-level languages like CUDA or OpenCL, the advent of high-level Python
libraries such as Numba~\citep{lam2015numba}, CuPy~\citep{okuta2017cupy} 
%(Okuta et al., 2017), 
and
PyCUDA/PyOpenCL~\citep{klockner2012pycuda}
% (Klöckner, 2012) 
has democratized access to GPU acceleration.
These tools offer user-friendly interfaces and just-in-time compilation,
allowing scientists to harness the power of GPUs within the familiar Python
ecosystem, thereby significantly accelerating scientific discovery and
broadening the adoption of accelerated computing across various research
domains.

% [JK: this is a super rough first narrative draft.]
% Machine learning, and in particular deep learning methods, have led to
% major breakthroughs in many different fields in recent years [CITES].
% GPUs have been a key enabling factor in this [details and cites].
% The PyData ecosystem [cites] has also enabled this. Techologies like
% [Cupy, lots of others] substantially lower the bar for writing efficient
% GPU code.
% Deep learning methods depend crucially on efficient access to vast
% datasets for training, and Zarr is emerging as a popular choice
% [cites from geosciences?]. The RAPIDs GPU Accelerated data science
% library [which does] now supports Zarr through [kvikio, more?].

@@

We first illustrate the ease with which computations can be
be peformed on the GPU, with little or no preprocessing of the
data. To do this we reimplemented the af-dist calculation
described above using CuPy~\citep{okuta2017cupy}, a GPU accelerated library
of core operations for scientific computing.
All of the operations required to compute  af-dist are
performed on the GPU, mostly with minimal changes required
(e.g.\ replacing \texttt{np.digitize} with \texttt{cp.digitize}).
We perform the core operation of genotype classification using the
CuPy JIT compiler to generate a custom CUDA kernel, which
is also very similar to the Numba compiled function used in other
benchmarks. In the simplest version, we iterate over the chunks
in the \texttt{call\_genotype} array (stored as 30 chunks, consuming
a total of 7.82MiB), decompress each to a local
NumPy array and then copying the data to GPU memory by creating
the corresponding CuPy array. For this version, the total time
in the genotype classification kernel was 25ms with an
overall data transfer time of 179ms (we discount decompression
time here; see next paragraph). For comparison,
our standard Numba compiled classification function required
a total of 1.46s running on a single CPU core.
Another approach in which we load the entire
\texttt{call\_genotype} array into GPU memory at once (possible
here as the $96514 \times 2504 \times 2$ array requires
460MiB) and uses
a more sophisticated kernel in which we map GPU resources to
subsets of the array in two dimensions,
reduces transfer
time to 138ms and processing time to 6ms.
